---
title: "Lab 5"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "C:/Users/timmy/OneDrive/Desktop/ESP106/lab5") #use this to set your working directory so knitr understands it
```

## Lab 5

### Due Tuesday Feb1st - Recommended to complete this before starting the midterm

This lab we will look at some data from the plastic trash piced up during clean-up events around the world. I took this dataset from the Tidy Tuesday website. You can read the documentation [here](https://github.com/rfordatascience/tidytuesday/blob/master/data/2021/2021-01-26/readme.md), including the references and description of the different column names.

I have done some pre-processing of the data for you for this lab, to create two more easy-to-use dataframes.

First read in the countrytotals.csv data frame
```{r echo=FALSE}
countrytotals = read.csv("countrytotals.csv")
as.data.frame(countrytotals)
str(countrytotals)

```

Have a look at the data frame. Then column "total" gives the total number of pieces of plastic picked up in that country in 2020. The columns "num_events" and "volunteers" give the number of trash pick-up events and the number of volunteers in that country. We are going to use this to investigate where the plastic trash problem is worst.

1. What 5 countries had the worst plastic problem as measured by the number of pieces of trash picked up?

```{r echo=FALSE}
order(countrytotals$total, decreasing = TRUE)
countrytotals[49,]
#row 35,37,46,21,49
#Nigeria, Phillipines, Switzerland, India, Togo
#hint order(datacolumn, decreasing=TRUE) will give you the ranking of the countries by the values in the data column
#you can use this to re-order the country data frame, ordering it from the country with the most trash picked up to the least
```

**Answer:** 
#row 35,37,46,21,49 had the highest total pieces of trash picked up, which corresponded to
#Nigeria, Phillipines, Switzerland, India, Togo in order

2. Make a plot showing the distribution of volunteers across coutnries

```{r echo = FALSE}
library(ggplot2)
ggplot(countrytotals, aes(x=countrytotals$country, y=countrytotals$volunteers)) + geom_bar(stat = "identity") + coord_flip() +
  theme(axis.text=element_text(size=5.5)) +
  labs(x="country", y="pieces of trash collected")
```

3. Notice that there is a lot of variation across countries in the number of volunteers involved in trash pickup. What problem might that cause for the interpretation of your answer to question 1?

**Answer:** 
#The number of volunteers directly correlate to the number of trashed picked up, therefore having a high amount of trashed picked up may not necessarily be representative of the amount of trash a country has but the number of volunteers they have or the size of their volunteer program

4. Add a column to the data frame creating a variable that should be more closely related to the presence of plastic pollution in the country

```{r echo=FALSE}

trashp_vol_pevent = countrytotals$total / countrytotals$volunteers / countrytotals$num_events
as.vector(trashp_vol_pevent)
str(countrytotals)
countrytotals = cbind(countrytotals, trashp_vol_pevent)
str(countrytotals)

```

5. What 5 countries have the worst plastic pollution, as measured by this new variable?

```{r echo=FALSE}
order(countrytotals$trashp_vol_pevent, decreasing = TRUE)
countrytotals[34,]
per_trash = na.omit(countrytotals[order(countrytotals$trashp_vol_pevent),])
#49,15,25,43,34
#Togo, El Salvador, Japan, South Africa, Netherlands
```

**Answer:** 
#I divided the pieces of trash picked up by the number of volunteers and the number of events, which is the number of trash per volunteer on average at an event. After making this new column within the data frame, the 5 countries with the most trash pollution according to this are Togo, El Salvador, Japan, South Africa, and the Netherlands

Now we will make a plot of the variation in the types of trash and how it differs around the world. Read in the continenttypes.csv data frame. This gives the breakdown of the different types of plastic collected on each continet in 2020 and the total number of pick up events.

```{r echo=FALSE}
continent = read.csv("continenttypes.csv")
continent

```

6. Add a column to this data frame with a variable that captures the existence of different types of plastic trash, controlling for the intensity of the pick-up effort in different continents

```{r echo=FALSE}
perptrash = continent$total / continent$events
continents = cbind(continent, perptrash)
```

7. Make a plot using ggplot showing both the total amount and distribution of types of plastic picked up in each continent in the average pick-up event.

Hint: Check out options in the [R graph gallery](https://www.r-graph-gallery.com)

```{r echo=FALSE}
library(ggplot2)
ggplot(continents, aes(fill=plastic_type, x=continents$continent, y=continents$perptrash)) +
  geom_bar(stat = "identity", position = "stack") +
labs(x = "continent", y = "pieces of plastic collected per event", fill = "plastic type")

```

8. Try uploading your R markdown file and plot to your Git Hub repository. Don't put your knitted HTML file in Github - these are large files that are not designed to be stored on Github

