---
title: "week3_lab2"
author: "Timothy Chen"
date: "2023-01-25"
output: html_document
---

Hello 


In this lab we will start by reading merging in data on economic development and indoor and outdoor air pollution. Then we will practice making some graphs with it.



~~1. First read in the three csv files: gdppercapitaandgini and airpollution!!~~

**#Setting up working directory and column names**
```{r}
setwd("C:/Users/timmy/OneDrive/Desktop/ESP106/lab3/ESP106_week3_lab1")
GDP = read.csv('gdppercapiandgini.csv')
air = read.csv('airpollution.csv')
colnames(GDP) = c('countries', 'code', 'year', 'population', 'continent', 'gini', 'GDP per capita')
colnames(air) = c('countries', 'code', 'year', 'particulate matter', 'household', 'ozone', 'air pollution')
colnames(GDP)
colnames(air)
```

**#calculating the total outdoor air pollution for Taiwan and China**
```{r}
totaldeaths = air$`particulate matter` + air$ozone
airpollution = cbind(air,totaldeaths)
tw = airpollution[airpollution$countries=="Taiwan",]
ch = subset(airpollution, countries=="China")
```

**#plot 1 of Taiwan and China's indoor and outdoor air pollution from 1985 to 2015**
```{r}
par(mfrow=c(1,1))
par(mar=c(5,4,2,2))
plot(tw$year, tw$household, ylim = c(0, 120),
     ylab = "Deaths per 100,000 people", 
     xlab ="Year", type = "l", 
     main = "Death rates from indoor and outdoor air pollution (1990-2015)", 
     las = 1, lwd = 3, col = "skyblue"
     )
lines(tw$year, tw$totaldeaths, type = 'l', lwd = 3, lty = 3, col = "skyblue")
lines(tw$year, ch$household, type = 'l', lwd = 3, col = "coral")
lines(tw$year, ch$totaldeaths, type = 'l', lwd = 3, lty = 3, col = 'coral')
legend("topright", c("Taiwan indoor", "Taiwan outdoor", "China indoor", "China outdoor"),
        col = c('skyblue','skyblue','coral','coral'), lty = c(1, 3, 1, 3), lwd = 2)
```

**#Merging GDP and air pollution data, by countires, country code, and year**  
```{r}
airGDP = merge(GDP, airpollution, by = c("countries", "code", "year"))
```

**#plot2 plotting log GDP against indoor and outdoor pollution death rates of available all countries**
```{r}
par(mfrow=c(1,2))
par(mar=c(5,4,4,2))
indoor = plot(log(airGDP$`GDP per capita`), airGDP$household, type = "p", pch=20, cex=0.5, 
     xlab = "log per Capita GDP", ylab = "deaths per 100,000 people", main = "Indoor air pollution",
     las = 1, col="seagreen", ylim = c(0,300))
out = plot(log10(airGDP$`GDP per capita`), airGDP$totaldeaths, type = "p", pch=20, cex=0.5,
       xlab = "log per Capita GDP", ylab = "deaths per 100,000 people", main = "Outdoor air pollution",
       las = 1, col = "orange1", ylim = c(0,300))

```


~~By bye~~